Point-process based representation learning for Electronic Health Records

### 2023-03-08 Manuscript_v1.tex
update

@@

% Table generated by Excel2LaTeX from sheet 'time concat'
\begin{table*}[htbp]
    \centering
    \caption{Add caption}
      \begin{tabular}{cccccccccccccc}
      \toprule
      \toprule
            &       & \multicolumn{8}{c}{TEEDAM}                                    &       &       &       &  \\
  \cmidrule{3-10}          &       & \multicolumn{2}{c}{AE (next mark)} &       & \multicolumn{2}{c}{PP(single+mark)} &       & \multicolumn{2}{c}{PP (MC/ML)} &       &       &       &  \\
  \cmidrule{3-4}\cmidrule{6-7}\cmidrule{9-10}    Dataset & Metric & concat & sum   &       & concat & sum   &       & concat & sum   & Latent & SAHP  & THP   & GRU-CP \\
      \midrule
      \multirow{2}[2]{*}{SO (MC)} & LL/\#events & \textit{ND} & \textit{ND} &       & -0.56 & -0.57 &       & -2.04 & -2.05 & -1.54 & -1.86 & -1.84 & \textit{NR} \\
            & F1-score & \textbf{38.46} & 36.91 &       & 36.04 & 34.91 &       & 32.65 & 31.67 & 28.34 & 24.12 & 23.89 & 26 \\
      \midrule
      \multirow{2}[2]{*}{ReTweet (MC)} & LL/\#events & \textit{ND} & \textit{ND} &       & -6.91 & -8.036 &       & -11   & -23.38 & -3.89 & -4.56 & -4.57 & \textit{NR} \\
            & F1-score & \textbf{62.69} & 61.39 &       & 53.91 & 38.63 &       & 53    & 35.17 & 58.29 & 53.92 & 53.86 & \textit{NR} \\
      \midrule
      \multirow{2}[2]{*}{Synthea (ML)} & LL/\#events & \textit{ND} & \textit{ND} &       & -2.405 & -2.599 &       & -7.254 & -6.58 & \textit{ND} & \textit{ND} & \textit{ND} & \textit{NR} \\
            & AUROC & \textbf{89.58} & 89.19 &       & 64.98 & 63.95 &       & 60.95 & 60.65 & \textit{ND} & \textit{ND} & \textit{ND} & 0.85(.014) \\
      \midrule
      \multirow{2}[2]{*}{ReTweet (ML)} & LL/\#events & \textit{ND} & \textit{ND} &       & 2.045 & 1.863 &       & -1.517 & -1.601 & \textit{ND} & \textit{ND} & \textit{ND} & \textit{NR} \\
            & AUROC & 69.63 & 68.54 &       & 69.31 & 68.05 &       & \textbf{74.22} & 71.27 & \textit{ND} & \textit{ND} & \textit{ND} & 0.611(0.001) \\
      \bottomrule
      \bottomrule
      \end{tabular}%
    \label{tab:1}%
  \end{table*}%

@@

Table \ref{tab:1} shows the performance metrics of NPP4EHR for different datasets in various scenarios. We can see that in most cases time concatenation leads to better results in terms of LL/\#events and F1-score/AUROC. While adding time encodings to the event embeddings is the default practice in the point process literature \cite*{zhangSelfAttentiveHawkesProcess2020,zuoTransformerHawkesProcess2020a} as well as natural language processing, we can see that time concatenation would allow model to achieve better results.

Another interesting fact is that the simple auto-encoder for next event type prediction achieves better results for SO,RT(MC) and SYN(ML) compared to the point process scenario. These datasets have been widely used in the point process literature to show the effectiveness of point process, however, we show that this simple baseline may perform better for some datasets. We remark again that point process loss functions (\ref*{eq:CIF-mc},\ref*{eq:CIF-ml},\ref*{eq:CIF-marked}) will reduce to the loss function of AE(next mark) if we neglect the integral term.

RT(MC) is the only dataset in which the point process with multi-label loss (equation \ref*{eq:CIF-ml}) performs better indicating the real advantage of modeling non-event likelihood of the point process in this dataset. For future research in the neural point process literature, we would recommend comparing model performance with this simple autoencoder model.

@@

Table \ref*{tab:2} shows the result for the utility of state encodings for the estimation of negative likelihood. We see that in all cases, state encoding leads to higher AUROC for the next event type prediction and LL/\#event as well. It is intuitive that in a hospital, the absolute value of patient states can be useful for ordering future laboratory events. We have evaluated our model in a healthcare database, however, it can be further evaluated on other event sequence data where additional information is available.

@@

Table \ref*{tab:3} indicates the result for Mortality/sepsis prediction task across different settings and hospital centers. Generally, we can see that in the mc1 setting, has the best performance because of the larger training dataset from other centers and the test center. In addition, event encoding module is useful for 2/5 cases while in 1/5 cases degrades the performance.

@@

\begin{figure*}[!t]
    \centerline{\includegraphics{images/tsne.png}}
    \caption{Magnetization as a function of applied field.
    It is good practice to explain the significance of the figure in the caption.}
    \label{fig1}
    \end{figure*}

Fig 1 visualizes the t-SNE plot for the learned embeddings of [] with and without event encodings. Besides, we visualize the sampling pattern of an example patient and its neighbors. As can be seen,  in TEEDAM neighbors have more similar pattern compared to the baseline. Although event encoding does not necessarily leads to better performance, we can learn a better representation that considers both absolute values and patterns.
The 10 nearest neighbor pattern similarity (10-nnps) of positive labels for TEE+DAM ($0.903 \pm 0.058 $) is signifcantly higher than (DAM) ($0.832 \pm 0.095$)($pvalue \leq 0.001$).
One application could be time series generation.

>>>>

% Table generated by Excel2LaTeX from sheet 'datasets'
\begin{table}[htbp]
    \centering
    \caption{Add caption}
      \begin{tabular}{ccccc}
      \toprule
      \toprule
      Dataset & Task  & \# classes & \# samples & Avg. length \\
      \midrule
      Stack Overflow & Multi-class & 22    & 6633  & 72 \\
      ReTweets & Multi-class & 3     & 24000 & 109 \\
      ReTweets & Multi-label & 3     & 20000 & 104 \\
      Synthea & Multi-label & 357   & 12000 & 43 \\
      Physionet 2012 & Multi-label & 25    & 12000 & 21 \\
      Physionet 2019 & Multi-label & 25    & 12000 & 21 \\
      \bottomrule
      \bottomrule
      \end{tabular}%
    \label{tab:data}%

  \end{table}%

@@

% Table generated by Excel2LaTeX from sheet 'time concat'
% Table generated by Excel2LaTeX from sheet 'time concat'
\begin{table*}[htbp]
    \centering
    \caption{Add caption}
      \begin{tabular}{ccccccccccccccc}
      \toprule
      \toprule
            &       & \multicolumn{8}{c}{TEEDAM}                                    &       &       &       &       &  \\
  \cmidrule{3-10}          &       & \multicolumn{2}{c}{AE (next-mark)} &       & \multicolumn{2}{c}{PP(single+mark)} &       & \multicolumn{2}{c}{PP (MC/ML)} &       & \multicolumn{4}{c}{Baselines} \\
  \cmidrule{3-4}\cmidrule{6-7}\cmidrule{9-10}\cmidrule{12-15}    Dataset & Metric & concat & sum   &       & concat & sum   &       & concat & sum   &       & Latent & SAHP  & THP   & GRU-CP \\
      \midrule
      \multirow{2}[2]{*}{SO (MC)} & LL/\#events & \textit{ND} & \textit{ND} &       & -0.56 & -0.57 &       & -2.04 & -2.05 &       & -1.54 & -1.86 & -1.84 & \textit{NR} \\
            & F1-score & \textbf{38.46} & 36.91 &       & 36.04 & 34.91 &       & 32.65 & 31.67 &       & 28.34 & 24.12 & 23.89 & 26 \\
      \midrule
      \multirow{2}[2]{*}{ReTweet (MC)} & LL/\#events & \textit{ND} & \textit{ND} &       & -6.91 & -8.036 &       & -11   & -23.38 &       & -3.89 & -4.56 & -4.57 & \textit{NR} \\
            & F1-score & \textbf{62.69} & 61.39 &       & 53.91 & 38.63 &       & 53    & 35.17 &       & 58.29 & 53.92 & 53.86 & \textit{NR} \\
      \midrule
      \multirow{2}[2]{*}{Synthea (ML)} & LL/\#events & \textit{ND} & \textit{ND} &       & -2.405 & -2.599 &       & -7.254 & -6.58 &       & \textit{ND} & \textit{ND} & \textit{ND} & \textit{NR} \\
            & AUROC & \textbf{89.58} & 89.19 &       & 64.98 & 63.95 &       & 60.95 & 60.65 &       & \textit{ND} & \textit{ND} & \textit{ND} & 0.85(.014) \\
      \midrule
      \multirow{2}[2]{*}{ReTweet (ML)} & LL/\#events & \textit{ND} & \textit{ND} &       & 2.045 & 1.863 &       & -1.517 & -1.601 &       & \textit{ND} & \textit{ND} & \textit{ND} & \textit{NR} \\
            & AUROC & 69.63 & 68.54 &       & 69.31 & 68.05 &       & \textbf{74.22} & 71.27 &       & \textit{ND} & \textit{ND} & \textit{ND} & 0.611(0.001) \\
      \bottomrule
      \bottomrule
      \end{tabular}%
    \label{tab:1}%
  \end{table*}%

@@

Table \ref{tab:1} shows the performance metrics of NPP4EHR for different datasets in various scenarios. We can see that in most cases time concatenation leads to better results in terms of LL/\#events and F1-score/AUROC. While adding time encodings to the event embeddings is the default practice in the point process literature \cite*{zhangSelfAttentiveHawkesProcess2020,zuoTransformerHawkesProcess2020a} as well as natural language processing, we can see that time concatenation would allow our model to achieve better results in terms of log-likelihood and next event type prediction.

Another interesting fact is that the simple auto-encoder for next event type prediction (AE(next-mark)) achieves better results for SO, RT(MC) and SYN(ML) compared to other point process models. These datasets have been widely used in the point process literature to show the effectiveness of NTPP models, however, we show that this simple baseline may perform better for some datasets. We remark again that point process loss functions (\ref*{eq:CIF-mc},\ref*{eq:CIF-ml},\ref*{eq:CIF-marked}) will reduce to the loss function of AE(next-mark) if we neglect the integral term.

RT(MC) is the only dataset in which the point process with multi-label loss (\ref*{eq:CIF-ml}) performs better indicating the real advantage of modeling the non-event likelihood of the point process in this dataset. For future research in the NTPP literature, we would recommend AE(next-mark) as a simple baseline for NTPP architectures.

@@

% Table \ref*{tab:2} shows the results for the utility of state encodings for the estimation of negative likelihood. We see that in all cases, state encoding leads to higher AUROC for the next event type prediction and LL/\#event as well. The noisy model (TEE+noise) does not improve performance reflecting the fact that the observed performance gain comes from unique information from patient's states.

The results for the utility of state encodings for the estimation of negative likelihood are presented in Table \ref*{tab:2}. Our analysis indicates that state encoding consistently results in higher AUROC for predicting the next event type and a higher LL/\#event ratio in all cases. Notably, incorporating noise into the model (TEE+noise) does not yield any improvements in performance. This finding suggests that the observed performance gain is attributable to unique information derived from patient states, rather than extraneous factors.

% It is also intuitive that in a hospital, the absolute value of patient states can be useful for ordering future laboratory events. We have evaluated our model on a healthcare database, however, it can be further evaluated on other event sequence data where additional information is available.

It is also intuitive that in a hospital setting, the absolute value of patient states could prove useful in determining the order of future laboratory events. While we have evaluated our model using a healthcare database, we believe that further assessment could be conducted on other event sequence data that includes additional information.

@@

Table \ref*{tab:3} indicates the result for the mortality/sepsis prediction task across different settings and hospital centers. Generally, we can see that the mc1 setting has the best performance because of the larger training dataset from other centers and the test center. In addition, event encoding module is useful for 2/5 cases while in 1/5 cases degrades the performance.

@@

\begin{figure*}[!t]
    \centerline{\includegraphics{images/tsne.png}}
    \caption{Magnetization as a function of applied field.
    It is good practice to explain the significance of the figure in the caption.}
    \label{fig2}
    \end{figure*}

We have visualized the learned patients' embeddings for an example dataset and setting (P12-seft). Fig. 1-a and b visualize the t-SNE plot for the learned representations of TEE+DAM and TEE, respectively.

The 10 nearest neighbor pattern similarity (10-nnps) of positive labels for TEE+DAM ($0.903 \pm 0.058 $) is significantly higher than (DAM) ($0.832 \pm 0.095$)($pvalue \leq 0.001$). The sampling pattern of an example patient and its 4 nearest neighbors (in the embedding space) is visualized in Fig. \ref*{fig2}-c and d for TEE+DAM and TEE, respectively. As can be seen, in TEE+DAM, the 4 neighbors have a more similar pattern compared to the example patient.

We argue that event encoding could lead to better representation learning in EHRs although it does not necessarily leads to better performance. Patient representation that can respect the irregular sampling pattern in EHRs could have various applications in synthetic data generation.
